---
title: "Working with link functions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with link functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 3.5
)
```

```{r setup}
library(linkfunctions7)
```

A link function maps a constrained parameter $\theta$ to an unconstrained
linear predictor $\eta$, so that a model can be fitted without fighting the
parameter's boundary. A variance must stay positive and a probability must
stay in $(0, 1)$; the log and logit links move the optimization onto the
whole real line and map the result back.

Unlike a pair of `log()` and `plogis()` calls, every link in
**linkfunctions7** carries its **exact analytical derivatives up to fifth
order**, in both directions, together with a diagnostic that verifies them.
A modeling package needs these derivatives: a Newton or Fisher-scoring step
uses the derivative of the inverse link, and a higher-order correction uses
the orders above it. The fifth order is needed by models whose predictor is
defined by a recursion, since each order of differentiation through the
recursion involves one more order of the link.

## A link is an object

Each link is created by a constructor and prints its name, domain, and any
parameters:

```{r}
log_link()
logit_link()
```

The two directions and their derivatives are generics that dispatch on the link:

```{r}
lk <- log_link()
theta <- c(0.5, 1, 2)

linkfun(lk, theta)       # eta = g(theta) = log(theta)
linkinv(lk, linkfun(lk, theta))   # back to theta
```

The forward derivatives are taken with respect to $\theta$, the inverse ones with
respect to $\eta$:

```{r}
eta <- linkfun(lk, theta)

dlinkfun(lk, theta)      # g'(theta)
dlinkinv(lk, eta)        # (g^{-1})'(eta)
```

These derivatives are closed-form expressions, not finite differences. For the
log link, $g(\theta) = \log\theta$ and $g'(\theta) = 1/\theta$:

```{r}
all.equal(dlinkfun(lk, theta), 1 / theta)
```

## Every order, either direction

`linkderiv()` and `linkinvderiv()` reach any order from zero to five without
naming a separate function:

```{r}
lk <- logit_link()
theta <- c(0.2, 0.5, 0.8)
eta <- linkfun(lk, theta)

sapply(0:5, function(k) linkinvderiv(lk, eta, order = k))
```

Order zero is the function itself; the columns are the successive derivatives of
the inverse logit at those three points.

The fifth forward derivative of the exponential links has a limited range.
These links floor $\theta = e^{\eta}$ at a value chosen so that the fourth
forward derivative $-6/\theta^4$ stays representable with a factor of four to
spare, and at that value the fifth, $24/\theta^5$, is not. The forward fifth derivative of a log,
lower- or upper-bounded link is therefore `Inf` below about $\eta = -141$,
while the fourth remains finite to $\eta = -200$. Raising the floor to cover
the fifth would clamp $\theta$ over a range where `linkinv()` is now exact, so
the floor is left where it is. In the lower tail the inverse direction, which
a chain rule onto the link scale uses, is finite at every order.

## Seeing a link

Every link has a `plot()` method that draws the forward and inverse
transformations over sensible ranges:

```{r}
plot(logit_link())
```

The **softplus** link illustrates the differences between positivity links.
Like the log it keeps $\theta$ positive, but for large $\eta$ it grows linearly
instead of exponentially, so a large linear predictor does not produce an
extremely large parameter:

```{r}
plot(softplus_link(a = 1))
```

## Bounded parameters

`bounded_link()` maps an interval to the whole real line. Give it a lower bound,
an upper bound, or both:

```{r}
lk <- bounded_link(lwr = -3, upr = 2)
eta <- seq(-3, 3, length.out = 5)
theta <- linkinv(lk, eta)
theta
```

`theta` stays inside $[-3, 2]$ whatever $\eta$ is, and the derivatives are
available here too:

```{r}
dlinkfun(lk, theta)
d2linkfun(lk, theta)
```

## The available links

```{r echo = FALSE}
knitr::kable(data.frame(
  Constructor = c("`identity_link()`", "`log_link()`", "`logit_link()`",
                  "`probit_link()`", "`cloglog_link()`", "`loglog_link()`",
                  "`cauchit_link()`", "`rhobit_link()`", "`sqrt_link()`",
                  "`inverse_link()`", "`inverse_sq_link()`", "`power_link(lambda)`",
                  "`softplus_link(a)`", "`bounded_link(lwr, upr)`"),
  Domain = c("$(-\\infty, \\infty)$", "$(0, \\infty)$", "$(0, 1)$", "$(0, 1)$",
             "$(0, 1)$", "$(0, 1)$", "$(0, 1)$", "$(-1, 1)$", "$(0, \\infty)$",
             "$(0, \\infty)$", "$(0, \\infty)$", "$(0, \\infty)$",
             "$(0, \\infty)$", "$(\\text{lwr}, \\text{upr})$")
))
```

## Validating a link

Because the derivatives are hand-written, the package ships a diagnostic that
checks them. `check_link()` confirms that the link inverts cleanly, that it is
strictly monotone, that the inverse function theorem
$g'(\theta)\,(g^{-1})'(\eta) = 1$ holds, and that every analytical derivative
matches a numerical one:

```{r}
invisible(check_link(logit_link()))
```

The diagnostic accepts any link and is most useful for a newly written one.

## Defining a new link

A new link is a subclass of `link` with methods for the two directions and
their five derivatives each, twelve methods in all. Only `linkfun()` and
`linkinv()` are compulsory: the base class supplies every derivative that a
link does not implement, by one stencil applied to the highest order that the
link does implement. The example below writes four orders in each direction
and leaves the fifth to that fallback. It is the **negative log-log** link,
$\eta = -\log(-\log\theta)$ on $(0, 1)$.

```{r}
NegLogLog <- S7::new_class("NegLogLog", parent = link)

S7::method(linkfun, NegLogLog)  <- function(x, theta) -log(-log(theta))
S7::method(linkinv, NegLogLog)  <- function(x, eta)   exp(-exp(-eta))

S7::method(dlinkfun,  NegLogLog) <- function(x, theta) -1 / (theta * log(theta))
S7::method(d2linkfun, NegLogLog) <- function(x, theta) (1 + log(theta)) / (theta * log(theta))^2
S7::method(d3linkfun, NegLogLog) <- function(x, theta) {
  l <- log(theta)
  -(2 * l^2 + 3 * l + 2) / (theta * l)^3
}
S7::method(d4linkfun, NegLogLog) <- function(x, theta) {
  l <- log(theta)
  (6 * l^3 + 11 * l^2 + 12 * l + 6) / (theta * l)^4
}

S7::method(dlinkinv,  NegLogLog) <- function(x, eta) exp(-eta - exp(-eta))
S7::method(d2linkinv, NegLogLog) <- function(x, eta) exp(-eta - exp(-eta)) * (exp(-eta) - 1)
S7::method(d3linkinv, NegLogLog) <- function(x, eta) {
  e <- exp(-eta)
  exp(-eta - e) * (e^2 - 3 * e + 1)
}
S7::method(d4linkinv, NegLogLog) <- function(x, eta) {
  e <- exp(-eta)
  exp(-eta - e) * (e^3 - 6 * e^2 + 7 * e - 1)
}

neglog <- NegLogLog(link_name = "neglog-log", link_bounds = c(0, 1), link_params = NULL)
```

The diagnostic verifies the hand-written derivatives:

```{r}
invisible(check_link(neglog))
```

Both derivative rows report that they passed to order four, with one order
numerical. The fifth order comes from the fallback and is not checked, since
it would be compared with a numerical differentiation of the order below it;
the two would be the same computation and would agree even for a wrong link.
`link_fallback_orders()` returns the same information as a list.

A line reporting `[FAILED]` marks a derivative that disagrees with its
numerical reference.

## Further reading

- `?check_link`: the full diagnostic and the meaning of each check.
- `?linkinvderiv`: derivatives of any order.
- The [distributions7](https://github.com/statmodels7/distributions7) package,
  where these links become the scale on which distributions are fitted.
