---
title: "Working with link functions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with link functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 3.5
)
```

```{r setup}
library(linkfunctions7)
```

A link function maps a constrained parameter $\theta$ to an unconstrained linear
predictor $\eta$, so that a model can be fitted without fighting the parameter's
boundary. A variance must stay positive, a probability must stay in $(0, 1)$;
the log and logit links let you optimise over the whole real line and map back.

What sets **linkfunctions7** apart from a couple of `log()` and `plogis()` calls
is that every link carries its **exact analytical derivatives up to fourth
order**, in both directions --- and a diagnostic that proves they are right.
That is what a modelling package needs: a Newton or Fisher-scoring step wants the
derivative of the inverse link, and a higher-order correction wants the ones
above it.

## A link is an object

Each link is created by a constructor and prints its name, domain, and any
parameters:

```{r}
log_link()
logit_link()
```

The two directions and their derivatives are generics that dispatch on the link:

```{r}
lk <- log_link()
theta <- c(0.5, 1, 2)

linkfun(lk, theta)       # eta = g(theta) = log(theta)
linkinv(lk, linkfun(lk, theta))   # back to theta
```

The forward derivatives are taken with respect to $\theta$, the inverse ones with
respect to $\eta$:

```{r}
eta <- linkfun(lk, theta)

dlinkfun(lk, theta)      # g'(theta)
dlinkinv(lk, eta)        # (g^{-1})'(eta)
```

They are exact, not finite differences. For the log link $g(\theta) = \log\theta$
so $g'(\theta) = 1/\theta$, and indeed:

```{r}
all.equal(dlinkfun(lk, theta), 1 / theta)
```

## Every order, either direction

`linkderiv()` and `linkinvderiv()` reach any order from zero to four without
naming a separate function:

```{r}
lk <- logit_link()
theta <- c(0.2, 0.5, 0.8)
eta <- linkfun(lk, theta)

sapply(0:4, function(k) linkinvderiv(lk, eta, order = k))
```

Order zero is the function itself; the columns are the successive derivatives of
the inverse logit at those three points.

## Seeing a link

Every link has a `plot()` method that draws the forward and inverse
transformations over sensible ranges:

```{r}
plot(logit_link())
```

The **softplus** link is a good one to look at. Like the log it keeps $\theta$
positive, but instead of the log's global exponential it bends smoothly into a
straight line for large $\eta$, which makes it better behaved when the linear
predictor wanders:

```{r}
plot(softplus_link(a = 1))
```

## Bounded parameters

`bounded_link()` maps an interval to the whole real line. Give it a lower bound,
an upper bound, or both:

```{r}
lk <- bounded_link(lwr = -3, upr = 2)
eta <- seq(-3, 3, length.out = 5)
theta <- linkinv(lk, eta)
theta
```

`theta` stays inside $[-3, 2]$ whatever $\eta$ is, and the derivatives are
available here too:

```{r}
dlinkfun(lk, theta)
d2linkfun(lk, theta)
```

## The links on offer

```{r echo = FALSE}
knitr::kable(data.frame(
  Constructor = c("`identity_link()`", "`log_link()`", "`logit_link()`",
                  "`probit_link()`", "`cloglog_link()`", "`loglog_link()`",
                  "`cauchit_link()`", "`rhobit_link()`", "`sqrt_link()`",
                  "`inverse_link()`", "`inverse_sq_link()`", "`power_link(lambda)`",
                  "`softplus_link(a)`", "`bounded_link(lwr, upr)`"),
  Domain = c("$(-\\infty, \\infty)$", "$(0, \\infty)$", "$(0, 1)$", "$(0, 1)$",
             "$(0, 1)$", "$(0, 1)$", "$(0, 1)$", "$(-1, 1)$", "$(0, \\infty)$",
             "$(0, \\infty)$", "$(0, \\infty)$", "$(0, \\infty)$",
             "$(0, \\infty)$", "$(\\text{lwr}, \\text{upr})$")
))
```

## Trust, but verify

Because the derivatives are hand-written, the package ships a diagnostic that
checks them. `check_link()` confirms that the link inverts cleanly, that it is
strictly monotone, that the inverse function theorem
$g'(\theta)\,(g^{-1})'(\eta) = 1$ holds, and that every analytical derivative
matches a numerical one:

```{r}
invisible(check_link(logit_link()))
```

Run it on any link before you rely on it --- and certainly on any link you write
yourself.

## Defining your own link

A new link is a subclass of `link` with its ten methods: the two directions and
their four derivatives each. The pattern is short; here is a complete one for the
**negative log-log** link, $\eta = -\log(-\log\theta)$ on $(0, 1)$.

```{r}
NegLogLog <- S7::new_class("NegLogLog", parent = link)

S7::method(linkfun, NegLogLog)  <- function(x, theta) -log(-log(theta))
S7::method(linkinv, NegLogLog)  <- function(x, eta)   exp(-exp(-eta))

S7::method(dlinkfun,  NegLogLog) <- function(x, theta) -1 / (theta * log(theta))
S7::method(d2linkfun, NegLogLog) <- function(x, theta) (1 + log(theta)) / (theta * log(theta))^2
S7::method(d3linkfun, NegLogLog) <- function(x, theta) {
  l <- log(theta)
  -(2 * l^2 + 3 * l + 2) / (theta * l)^3
}
S7::method(d4linkfun, NegLogLog) <- function(x, theta) {
  l <- log(theta)
  (6 * l^3 + 11 * l^2 + 12 * l + 6) / (theta * l)^4
}

S7::method(dlinkinv,  NegLogLog) <- function(x, eta) exp(-eta - exp(-eta))
S7::method(d2linkinv, NegLogLog) <- function(x, eta) exp(-eta - exp(-eta)) * (exp(-eta) - 1)
S7::method(d3linkinv, NegLogLog) <- function(x, eta) {
  e <- exp(-eta)
  exp(-eta - e) * (e^2 - 3 * e + 1)
}
S7::method(d4linkinv, NegLogLog) <- function(x, eta) {
  e <- exp(-eta)
  exp(-eta - e) * (e^3 - 6 * e^2 + 7 * e - 1)
}

neglog <- NegLogLog(link_name = "neglog-log", link_bounds = c(0, 1), link_params = NULL)
```

Did the derivatives come out right? Ask the diagnostic rather than trusting the
algebra:

```{r}
invisible(check_link(neglog))
```

If any line came back `[FAILED]`, that is a derivative to revisit --- which is
exactly the mistake `check_link()` exists to catch.

## Where to go next

- `?check_link` --- the full diagnostic and what each check means.
- `?linkinvderiv` --- reaching any derivative order.
- The [distributions7](https://github.com/statmodels7/distributions7) package,
  where these links become the scale on which distributions are fitted.
